Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-04-06 ThesisJP/Background/background.tex
Changed from subfigure to subfloat in Backgroun chapter.
Small update. Started on concluding remarks section in Background chapter.

\begin{figure}
\centering
\subfigure[Kernel]{
	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}\label{fig:gpu_kernel}
}
\subfigure[Threads, blocks and grid]{
	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}\label{fig:gpu_grid}
}
\caption{The figure depicts how GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function. In this example the kernel function performs an in-place element-wise matrix square.}
\label{fig:gpu_grid_kernel}
\end{figure}

@@

\begin{figure}
\centering
\subfigure[]{
	\includegraphics[width=0.47\textwidth]{img/Diagram_of_the_human_heart.png}\label{fig:human_heart}
}
\subfigure[]{
	\includegraphics[width=0.47\textwidth]{img/ultrasound_grid.pdf}\label{fig:us_grid}
}
\caption{a) Overview of the human heart. Illustration from wikipedia.org. b) Illustration of probe geometry and layout for phased array imaging.}
\end{figure}

@@

\subsection{Beamforming}
If time delays are applied on the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. A cardiac probe is a phased probe, which means that the image is constructed from image lines which are gradually steered from $-\theta/2$ to $\theta/2$ radians. As depicted in Fig.\,\ref{fig:us_grid}, this results in an imaging sector which is $\theta$ radians wide. This is known as a 2D B-mode image (B for brightness). If image lines are acquired for multiple B-mode images which are tilted in the $yz$-plane we get what is known as a volumetric B-mode or a 3D ultrasound image. A probe for 3D imaging has several thousand elements arranged as a matrix in order to facilitate steering in both the $xz$ and $yz$-plane. In this thesis we will focus on 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{img/beamforming_das.eps}
\caption{Delay-and-sum array beamforming.}
\label{fig:das}
\end{figure}

The standard form of beamforming used on receive in ultrasound imaging is known as \textit{delay-and-sum} (DAS) beamforming: 
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output image line in the direction given by $\Delta_m[n]$.  From now on we will omit the dependency on $\Delta_m[n]$ and $x_m$ will refer to delayed element data. A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das}. If $p_o$ is the phase center of our array and $p$ is the point to which we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays has to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse. Transmit delays therefore do not vary within an image line.

@@

\subsection{GPUs in medical ultrasound imaging}
Ultrasound scanners generate an incredible amount of data. A modern system capable of 3D imaging will generate several tera bytes of data inside the probe handle during a clinical scan. Fortunately the data rate is reduced to the level of 2D imaging before the data reaches the scanner. The data rate of a 2D scan is "only" in order of giga bytes per second. To handle this amount of data, ultrasound imaging systems have, for the last twenty years, relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline \cite{Thomenius}. However, as the computational power of generic processors have increased, ASICs have gradually been replaced by software \cite{Guracar2013}. One of  the last remaining ASICs in ultrasound scanners is used for beamforming \cite{Thomenius}. As seen from (\ref{eq:das}), beamforming involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe. The input data rate is therefore very high. Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}. This is the same technique as reported implemented on a GPU by Yiu \textit{et al.} \cite{Yiu2011}. We therfore can conclude that beamforming is finally moving into software too. The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}. Software beamforming opens up for new possibilities for advanced processing of ultrasound data. One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), and which is introduce in the next section. The increased computational power introduced by GPUs makes it also possible to include more advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.

@@

\begin{figure}[t!]
\subfigure[Uniform weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}\label{fig:das_weights}
}
\subfigure[Capon weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}\label{fig:capon_weights}
}
\caption{Examples of array beampatterns where the array weights are either uniform (a) or derived using Capon's method (b). An interfering source is located at 20\degree. Notice the reduced sidelobe level for the Capon beampattern in the direction of the interfering source. Image courtesy of Jo Inge Buskenes.}
\label{fig:weights}
\end{figure}

The Capon beamformer produces a weight vector, as introduced in (\ref{eq:das}), based on the impinging signal and a constrained optimization problem. The problem is as follow \cite{Capon1969}:
\begin{align}
&\min_{\vec{w}} E\{|z[n]|^2\} \rightarrow \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria} \\
&\text{subjected to } \vec{w}^H\vec{a} = 1,
\end{align}
where \nom{$\mat{\hat{R}}$}{Sample covariance matrix (with regularization)} is a sample covariance matrix, and \nom{$(.)^H$}{Complex conjugate transpose} is the complex conjugate transpose. Hence, the resulting weight vector minimizes the output power (the criterion) while maintaining unit gain in the steering direction $\vec{a}$ (distortionless response). The benefit of Capon beamforming is clearly seen from Fig\,\ref{fig:weights} which present a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer. In this simulated case an interfering source is located at 20\degree{} which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated with around $13$ dB. Since the Capon weights are data-dependent the method "knows" where the interfering source is located and is able to position a zero in the direction of it (Fig.\,\ref{fig:capon_weights}). Hence, the interfering source will be highly attenuated independent of its direction. Another interesting case is when the interfering source is located within the mainlobe of the array beampattern. In theory a signal will be regarded by the Capon beamformer as an interfering source as long as the signal propagation vector does not match the steering vector ($\vec{a}$) exactly. If it is not possible to position a zero in the direction of the interferer the mainlobe will get shifted and scaled if this is whats needed in order to achieve minimum output power. Adaptive positioning of zeros and mainlobe shifting gives rice to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, and the positive implications of this effect \cite{Synnevag2009} is the main reason why this beamformer has been highly studied in the field of medical ultrasound imaging the last decade.

To "know" the direction of interfering sources the Capon beamformer has to estimate the statistics of the impinging wave field. This is done by estimating a sample covariance matrix, which for an active broadband system can be estimated in the following way \cite{Synnevag2009}:
\begin{align}
\mat{\breve{R}}[n] &= \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H\\
\vec{x}_l[n'] &= [x_l[n], \dotso, x_{l+L}[n]]
\end{align}
where\nomi{$K$}{Temporal averaging parameter (Capon)} $N_K = 2K + 1$ \nomi{$N_K$}{Number of samples used for temporal averaging (Capon)} is the number of samples used for temporal smoothing,\nomi{$L$}{Subarray length (Capon)} $\vec{x}_l$ is a $L$ long subarray, and $N_L = M-L+1$\nomi{$N_L$}{Number of subarrays used for spatial averaging (Capon)} is the number of subarrays used for spatial averaging.  Temporal averaging is used to obtain DAS-like speckle statistics in combination with large subarrays \cite{Synnevag2007a}, and subarray averaging is applied to reduce signal cancellation caused by coherent signals \cite{Reddy1987}. The number of temporal samples is usually selected relative to the pulse length. To ensure robustness against model errors and numerical stability the matrix is further loaded with a diagonal factor, 
\begin{align}
\epsilon &= d*\text{trace}\{\mat{\breve{R}}\}/L\\
\mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
\end{align} 
Hence, a white noise component proportional to the in-coherent signal power is added to the estimate \cite{Featherstone1997b}.

The solution to the constrained minimization problem in (\ref{eq:capon_optimization_criteria}) is, by the method of Lagrange multipliers \cite{VanTrees2003}, found to be
\begin{align}\label{eq:capon_weights}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L.
\end{align}
Since $\vec{w}$ is time (or range) dependent one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system. This is indeed computationally demanding, something which becomes clear when each step of the beamformer is examined closer. It is then found that constructing the covariance matrix has a computationally complexity of $O(L^2N_LN_K)$\nomi{O(.)}{Big O notation (algorithm complexity)}, inverting it has a computationally complexity of $O(L^3)$, and the final normalization and calculation of $\vec{w}$ is $O(L)$. For $L=M/2$, a subarray size often used in the literature, the first two steps both have a complexity of $O(L^3)$. These steps should therefore receive the same amount of attention when a real-time implementation of Capon beamforming is attempted.

A rough estimate of the number of FLOPS required by a Capon beamformer for real-time cardiac ultrasound imaging is then given by:
\begin{align}
\text{FLOPS}_{\text{Capon}} = L^3B_wf_c.
\end{align}
Which for our example, with $L=M/2=64$, sums up to a required throughput of more than $1/2$ TFLOPS. Even though this is just 10\% of the peak throughput of modern GPUs remember that the latter are only theoretical numbers, and that the actual throughput is highly algorithm dependent. The required throughput for Capon beamforming might therefore still be hard to achieve in practice \cite[\textbf{Paper II}]{So2011}. As for DAS beamforming this number also scales with higher relative bandwidth, pulse frequency, and more array elements since the subarray size typically is proportional to the number of array elements.

@@

To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging several simplified version have been proposed in the literature \cite{Asl2012, Kim}. Among the most interesting proposals we find the low complexity adaptive (LCA) beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}. In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data is located in prior to scan conversion. In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, hence, the spatial Fourier transform of the element data: 
\begin{align}
\vec{x}_{\text{BS},l} &= \mat{B}[x_l[n],\,\dotso,\,x_{l+L}[n]]^T\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{L}}e^{-j 2 \pi p q / L}.
\end{align}
The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$. When investigating the complexity of the BS-Capon beamformer (\textbf{Paper II}) we find that is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of selected beams.

@@

Both the LCA and BS-Capon beamformer is based upon a detailed investigation of how Capon beamfoming behaves when applied to medical ultrasound images. For the LCA beamformer, the windows selected by the Capon beamformer where investigated and a special set of predefined windows where defined. If one look closer at the windows selected for the LCA beamformer we see that half the set performs micro-steering of the mainlobe and the other half provides different mainlobe widths. The results from applying the LCA beamformer shows that the main benefit of applying Capon beamforming to medical ultrasound imaging is micro-steering of the mainlobe, and not attenuation of far off-axis backscatter.

The Capon beamformer is derived for narrowband applications and has traditionally been applied on medical ultrasound imaging "as is". Medical ultrasound imaging is, however, a broadband imaging modality and one could think that a broadband Capon beamformer therefore where strictly needed \cite{Holfort2009}. The investigation performed in \cite{Nilsen2009} showed that with pre-delayed beams, energy will be centered on-axis and as little as three near-axis beams is therefore adequate in order to achieve similar performance for BS-Capon and Capon beamforming. The two-way beam will be wider for unfocused (plane wave) transmit beams \cite{Holfort2009, Austeng2011}, but steering and focusing on receive will still make the main amount of energy arrive on axis. As shown in \textbf{Paper III} it is also possible to micro-steer with phase delays (narrowband) instead of time delays (broadband) without introducing large errors. Since Capon beamforming mainly performs main-lobe shifts smaller than the beam width, the difference between a broadband and narrowband Capon beamformer will be minimal \cite{Diamantis2014}. Broadband Capon beamforming is also significantly more computationally demanding than narrowband Capon beamforming.

@@

\section{Volume rendering}\label{sec:volren}
\begin{figure}[t!]
\centering
\subfigure[Ray-casting]{
	\includegraphics[width=0.4\textwidth]{img/Volumeraycasting.png}\label{fig:vr_rc}
}
\subfigure[Opacity transfer function]{
	\includegraphics[width=0.5\textwidth]{img/otf.png}\label{fig:vr_otf}
}
\caption{The principle behind volume ray casting. a) A ray is casted out from every pixel in the output image. Along the ray, samples are gathered from the volume. The samples are assigned a color and an opacity before they are combined into one pixel value. b) The opacity transfer function used to determine how opaque a sample should be.}
\label{fig:vr}
\end{figure}
A 3D ultrasound dataset is a cloud of scalar values that somehow needs to be presented to the user. One approach used for cardiac ultrasound  is to extract slices corresponding to different long-axis and short-axis views of the cardiac chambers. Another way is to project the whole volume into a two dimensional image. The latter technique is known as volume rendering, and it present for the user the three dimensional structure of the heart in one two-dimensional image. Many different methods for volume rendering exists \cite{brodlie2001}, and in this thesis we will focus on ray casting \cite{Levoy1988}.

@@

>>>>

\begin{figure}
\centering
\subfloat[Kernel]{
	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}\label{fig:gpu_kernel}
}
\subfloat[Threads, blocks and grid]{
	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}\label{fig:gpu_grid}
}
\caption{The figure depicts how GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function. In this example the kernel function performs an in-place element-wise matrix square.}
\label{fig:gpu_grid_kernel}
\end{figure}

@@

\begin{figure}
\centering
\subfloat[]{
	\includegraphics[width=0.47\textwidth]{img/Diagram_of_the_human_heart.png}\label{fig:human_heart}
}
\subfloat[]{
	\includegraphics[width=0.47\textwidth]{img/ultrasound_grid.pdf}\label{fig:us_grid}
}
\caption{a) Overview of the human heart. Illustration from wikipedia.org. b) Illustration of probe geometry and layout for phased array imaging.}
\end{figure}

@@

\subsection{Beamforming}
If time delays are applied on the signal emitted by the elements it is possible to steer and focus the ultrasound beam in space (Fig.\,\ref{fig:das_background}). This technique is known as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. A cardiac probe is a phased probe, which means that the image is constructed from image lines which are gradually steered from $-\theta/2$ to $\theta/2$ radians. As depicted in Fig.\,\ref{fig:us_grid}, this results in an imaging sector which is $\theta$ radians wide. This is known as a 2D B-mode image (B for brightness). If image lines are acquired for multiple B-mode images which are tilted in the $yz$-plane we get what is known as a volumetric B-mode or a 3D ultrasound image. A probe for 3D imaging has several thousand elements arranged as a matrix in order to facilitate steering in both the $xz$ and $yz$-plane. In this thesis we will focus on 2D imaging when the topic is adaptive beamforming, and 3D imaging when the topic is adaptive visualization.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{img/beamforming_das.eps}
\caption{Delay-and-sum array beamforming.}
\label{fig:das_background}
\end{figure}

The standard form of beamforming used on receive in ultrasound imaging is known as \textit{delay-and-sum} (DAS) beamforming: 
\begin{align}\label{eq:das_background}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
Here, \nom{$M$}{Number of array elements} is the number of elements in the array, $x_m$ is the per-element data, \nom{$\Delta_m$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, \nom{$\vec{w}$}{Array weight vector (apodization)} is the array \textit{weight vector} or \textit{apodization}, $\vec{x}$ is a pre-delayed vector of element data, and $z[n]$ is the output image line in the direction given by $\Delta_m[n]$.  From now on we will omit the dependency on $\Delta_m[n]$ and $x_m$ will refer to delayed element data. A schematic overview of delay-and-sum array beamforming is depicted in Fig.\,\ref{fig:das_background}. If $p_o$ is the phase center of our array and $p$ is the point to which we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_o+p| - |p_m+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays has to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse. Transmit delays therefore do not vary within an image line.

@@

\subsection{GPUs in medical ultrasound imaging}
Ultrasound scanners generate an incredible amount of data. A modern system capable of 3D imaging will generate several tera bytes of data inside the probe handle during a clinical scan. Fortunately the data rate is reduced to the level of 2D imaging before the data reaches the scanner. The data rate of a 2D scan is "only" in order of giga bytes per second. To handle this amount of data, ultrasound imaging systems have, for the last twenty years, relied heavily on application-specific integrated circuits (\nom{ASIC}{Application-specific integrated circuit}) for all steps in the processing pipeline \cite{Thomenius}. However, as the computational power of generic processors have increased, ASICs have gradually been replaced by software \cite{Guracar2013}. One of  the last remaining ASICs in ultrasound scanners is used for beamforming \cite{Thomenius}. As seen from (\ref{eq:das_background}), beamforming involves a reduction of the data rate with a factor equal to the number of elements in the ultrasound probe. The input data rate is therefore very high. Nevertheless, real-time beamforming has recently been achieved on a GPU \cite{Song2012}, and a French company (SuperSonic Imagine) are utilizing GPUs in their plane wave beamforming algorithm \cite{Tanter2014}. This is the same technique as reported implemented on a GPU by Yiu \textit{et al.} \cite{Yiu2011}. We therfore can conclude that beamforming is finally moving into software too. The largest players in the medical ultrasound business, General Electric and Philips, are also well aware of where the trend is going \cite{Thomenius2012} \cite{Metz2011}. Software beamforming opens up for new possibilities for advanced processing of ultrasound data. One example is Capon beamforming which is explored in this thesis (\textbf{Paper\,I} and \textbf{II}), and which is introduce in the next section. The increased computational power introduced by GPUs makes it also possible to include more advanced 3D visualization as presented in \cite{solteszova2010multidirectional}, \textbf{Paper\,IV}, and \textbf{IX}.

@@

\begin{figure}[t!]
\subfloat[Uniform weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}\label{fig:das_weights}
}
\subfloat[Capon weights]{
	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}\label{fig:capon_weights}
}
\caption{Examples of array beampatterns where the array weights are either uniform (a) or derived using Capon's method (b). An interfering source is located at 20\degree. Notice the reduced sidelobe level for the Capon beampattern in the direction of the interfering source. Image courtesy of Jo Inge Buskenes.}
\label{fig:weights}
\end{figure}

The Capon beamformer produces a weight vector, as introduced in (\ref{eq:das_background}), based on the impinging signal and a constrained optimization problem. The problem is as follow \cite{Capon1969}:
\begin{align}
&\min_{\vec{w}} E\{|z[n]|^2\} \rightarrow \min_{\vec{w}} \vec{w}^H \mat{\hat{R}} \vec{w} \label{eq:capon_optimization_criteria_background} \\
&\text{subjected to } \vec{w}^H\vec{a} = 1,
\end{align}
where \nom{$\mat{\hat{R}}$}{Sample covariance matrix (with regularization)} is a sample covariance matrix, and \nom{$(.)^H$}{Complex conjugate transpose} is the complex conjugate transpose. Hence, the resulting weight vector minimizes the output power (the criterion) while maintaining unit gain in the steering direction $\vec{a}$ (distortionless response). The benefit of Capon beamforming is clearly seen from Fig\,\ref{fig:weights} which present a sensitivity map (an \textit{array beampattern}) for DAS beamforming with uniform weights and the Capon beamformer. In this simulated case an interfering source is located at 20\degree{} which for the DAS beamformer with uniform weights (\ref{fig:das_weights}) will leak into the beamformer output with its power attenuated with around $13$ dB. Since the Capon weights are data-dependent the method "knows" where the interfering source is located and is able to position a zero in the direction of it (Fig.\,\ref{fig:capon_weights}). Hence, the interfering source will be highly attenuated independent of its direction. Another interesting case is when the interfering source is located within the mainlobe of the array beampattern. In theory a signal will be regarded by the Capon beamformer as an interfering source as long as the signal propagation vector does not match the steering vector ($\vec{a}$) exactly. If it is not possible to position a zero in the direction of the interferer the mainlobe will get shifted and scaled if this is whats needed in order to achieve minimum output power. Adaptive positioning of zeros and mainlobe shifting gives rice to the super-resolution capabilities of the Capon beamformer \cite{Synnevag2007}, and the positive implications of this effect \cite{Synnevag2009} is the main reason why this beamformer has been highly studied in the field of medical ultrasound imaging the last decade.

To "know" the direction of interfering sources the Capon beamformer has to estimate the statistics of the impinging wave field. This is done by estimating a sample covariance matrix, which for an active broadband system can be estimated in the following way \cite{Synnevag2009}:
\begin{align}
\mat{\breve{R}}[n] &= \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H\\
\vec{x}_l[n'] &= [x_l[n'], \dotso, x_{l+L}[n']]
\end{align}
where\nomi{$K$}{Temporal averaging parameter (Capon)} $N_K = 2K + 1$ \nomi{$N_K$}{Number of samples used for temporal averaging (Capon)} is the number of samples used for temporal smoothing,\nomi{$L$}{Subarray length (Capon)} $\vec{x}_l$ is a $L$ long subarray, and $N_L = M-L+1$\nomi{$N_L$}{Number of subarrays used for spatial averaging (Capon)} is the number of subarrays used for spatial averaging.  Temporal averaging is used to obtain DAS-like speckle statistics in combination with large subarrays \cite{Synnevag2007a}, and subarray averaging is applied to reduce signal cancellation caused by coherent signals \cite{Reddy1987}. The number of temporal samples is usually selected relative to the pulse length. To ensure robustness against model errors and numerical stability the matrix is further loaded with a diagonal factor, 
\begin{align}
\epsilon &= d*\text{trace}\{\mat{\breve{R}}\}/L\\
\mat{\hat{R}} &= \mat{\breve{R}} + \epsilon\mat{I}.
\end{align} 
Hence, a white noise component proportional to the in-coherent signal power is added to the estimate \cite{Featherstone1997b}.

The solution to the constrained minimization problem in (\ref{eq:capon_optimization_criteria_background}) is, by the method of Lagrange multipliers \cite{VanTrees2003}, found to be
\begin{align}\label{eq:capon_weights}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} \in \mathbb{C}^L.
\end{align}
Since $\vec{w}$ is time (or range) dependent one matrix has to be constructed and inverted for each data vector $\vec{x}[n]$ received by the system. This is indeed computationally demanding, something which becomes clear when each step of the beamformer is examined closer. It is then found that constructing the covariance matrix has a computationally complexity of $O(L^2N_LN_K)$\nomi{O(.)}{Big O notation (algorithm complexity)}, inverting it has a computationally complexity of $O(L^3)$, and the final normalization and calculation of $\vec{w}$ is $O(L)$. For $L=M/2$, a subarray size often used in the literature, the first two steps both have a complexity of $O(L^3)$. These steps should therefore receive the same amount of attention when a real-time implementation of Capon beamforming is attempted.

A rough estimate of the number of FLOPS required by a Capon beamformer for real-time cardiac ultrasound imaging is then given by:
\begin{align}
\text{FLOPS}_{\text{Capon}} = L^3B_wf_c.
\end{align}
Which for our example, with $L=M/2=64$, sums up to a required throughput of more than $1/2$ TFLOPS. Even though this is just 10\% of the peak throughput of modern GPUs remember that the latter are only theoretical numbers, and that the actual throughput is highly algorithm dependent. The required throughput for Capon beamforming might therefore still be hard to achieve in practice \cite[\textbf{Paper II}]{So2011}. As for DAS beamforming this number also scales with higher relative bandwidth, pulse frequency, and more array elements since the subarray size is selected proportional to the number of array elements.

@@

To reduce the processing requirement of the Capon beamformer for medical ultrasound imaging several simplified version have been proposed in the literature \cite{Asl2012, Kim}. Among the most interesting proposals we find the low complexity adaptive (LCA) beamformer by Synnev\aa{}g \textit{et al.} \cite{Synnevag2011} and the beamspace (BS) Capon beamformer by Nilsen and Hafizovic \cite{Nilsen2009}. In medical ultrasound imaging beamspace typically refers to the polar grid that cardiac ultrasound data is located in prior to scan conversion. In combination with the Capon beamformer, beamspace refers to the wavenumber-space representation of the recorded signals, hence, the spatial Fourier transform of the element data: 
\begin{align}
\vec{x}_{\text{BS},l} &= \mat{B}[x_l[n],\,\dotso,\,x_{l+L}[n]]^T\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{L}}e^{-j 2 \pi p q / L}.
\end{align}
The idea behind BS-Capon beamforming is that spatial beams with little information can be removed from $\mat{\hat{R}}$ by removing rows in $B$. When investigating the complexity of the BS-Capon beamformer (\textbf{Paper II}) we find that it is dominated by the transformation of data into beamspace, $O(N_LLN_b)$, where \nom{$N_b$}{Number of selected beams in beamspace} is the number of selected beams.

@@

\subsection{Narrowband vs Broadband Capon beamforming}
Both the LCA and BS-Capon beamformer is based upon a detailed investigation of how Capon beamfoming behaves when applied to medical ultrasound images. For the LCA beamformer, the windows selected by the Capon beamformer where investigated and a special set of predefined windows where defined. If one look closer at the windows selected for the LCA beamformer we see that half the set performs micro-steering of the mainlobe and the other half provides different mainlobe widths. The results from applying the LCA beamformer shows that the main benefit of applying Capon beamforming to medical ultrasound imaging is micro-steering of the mainlobe, and not attenuation of far off-axis backscatter.

The Capon beamformer is derived for narrowband applications and has traditionally been applied to medical ultrasound imaging "as is". Medical ultrasound imaging is, however, a broadband imaging modality and one could think that a broadband Capon beamformer therefore where strictly needed \cite{Holfort2009}. The investigation performed in \cite{Nilsen2009} showed that with pre-delayed data, energy will be centered on-axis and as little as three near-axis beams is therefore adequate in order to achieve similar performance for BS-Capon and Capon beamforming. The two-way beam will be wider for unfocused (plane wave) transmit beams \cite{Holfort2009, Austeng2011}, but steering and focusing on receive will still make the main amount of energy arrive on axis. As shown in \textbf{Paper III} it is also possible to micro-steer with phase delays (narrowband) instead of time delays (broadband) without introducing large errors. Since Capon beamforming mainly performs main-lobe shifts smaller than the beam width, the difference between a broadband and narrowband Capon beamformer will be minimal \cite{Diamantis2014}. Broadband Capon beamforming is also significantly more computationally demanding than narrowband Capon beamforming.

@@

\section{Volume rendering}\label{sec:volren}
\begin{figure}[t!]
\centering
\subfloat[Ray-casting]{
	\includegraphics[width=0.4\textwidth]{img/Volumeraycasting.png}\label{fig:vr_rc}
}
\subfloat[Opacity transfer function]{
	\includegraphics[width=0.5\textwidth]{img/otf.png}\label{fig:vr_otf}
}
\caption{The principle behind volume ray casting. a) A ray is casted out from every pixel in the output image. Along the ray, samples are gathered from the volume. The samples are assigned a color and an opacity before they are combined into one pixel value. b) The opacity transfer function used to determine how opaque a sample should be.}
\label{fig:vr}
\end{figure}
A 3D ultrasound dataset is a cloud of scalar values that somehow needs to be presented to the user. One approach used for cardiac ultrasound  is to extract slices corresponding to different long-axis and short-axis views of the cardiac chambers. Another way is to project the whole volume into a two dimensional image. The latter technique is known as volume rendering, and it present for the user the three dimensional structure of the heart in one two-dimensional image. Many different methods for volume rendering exists \cite{brodlie2001}, and in this thesis we will focus on ray casting \cite{Levoy1988}.

@@

\section{Concluding remarks}
All the presented problems in this chapter has in common that they require a processing throughput from ten to several hundred GFLOPS. Comparing this to the typical clock frequency of modern CPUs and GPUs (1.5 - 3 GHz) we see that GFLOPS problems requires parallel execution in order to achieve real-time performance.
